---
title: "Analysis of Break Free From Plastic Data with Supplemental Data"
format:
  html:
    theme: minty
    code-fold: true
    code-summary: "Show code"
    toc: true
    number-sections: true
    embed-resources: true
editor: visual
execute:
  warning: false
  message: false
---

```{r}

library(tidyverse)
library(httr)
library(jsonlite)
library(tidyjson)
library(glue)
library(countrycode)
library(gt)
library(hrbrthemes)
library(viridis)
library(hexbin)
library(patchwork)
```

```{r}
plastics <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2021/2021-01-26/plastics.csv')
```

# Data Context:

## Plastic Waste Data:

This data set was independently gathered by an individual names Sarah Sauve, in order to contribute to the global audit that the Break Free from Plastic organization holds each year. They collected data in their own town of St. John in Newfoundland from the year 2019 to 2020. The data set has information on the amount of different plastic waste (polyester, PVC, polyethylene, etc...) as well as the grand total of all plastic types, number of counting events, and number of volunteers at each cleanup. It is not specified how the data was collected.

## Population and GDP by Country Data:

Through the use of the World Bank Group's API, we were able to collect population and GDP data for each country in our original Plastic Waste data set between the years 2019 and 2020. The World Bank Group does not necessarily collect the data themselves, but relies on outside resources like countries National Statistic Agencies, Central Banks, and Government Agencies. Although the World Bank does not collect the data themselves, they coordinate and compile the data into easily accessible APIs to be used. The data is collected through multiple sources like household surveys, digital records, censuses, and secondary sources previously discussed. This data is collected in order to have a record of important statistics on countries that is easily accessible and can be analyzed at any given moment.

## CO~2~ Data:

Our final data set contains information about CO~2~ Emissions throughout history that was published by *Our World in Data*. Although *Our World in Data* published the data, they rely on outside sources such as the Global Carbon Project and other scientific databases to collect the data. The data set contains country-level CO~2~ emissions spanning as early as 1750 to 2024. However, to match our original data set, we restricted the years to 2019 and 2020. In order to actually come up with the data, the original collectors had to get creative. To estimate the CO~2~ emission for one country in any given year they would gather information on how much energy sources a country extracted in that year, then add the amount of fuel that country imported, and finally subtract any fuel the country exported. What was left is an estimate of the net consumption of energy sources in that year. They can then break down the energy sources into smaller categories and multiply the quantity of one fuel source by the emission factor to figure out how much CO~2~ was emitted by that energy source. They would then do that for all energy sources and sum them up to find the total amount of CO~2~ emission by a country in a given year. This may seem like a lot of work, but this is very important data. CO~2~ emissions are detrimental to the health and safety of Earth and its future. We must be tracking and analyzing how our CO~2~ emissions change year by year in order to better understand how we can mitigate the effect it will have on our future.

# Supplemental Data:

## API Data:

To better understand the difference of plastic waste in countries, we believed it would be important to factor in their population and GDP. Our justification for this was that with larger populations we are likely to see more plastic waste as there are more opportunities for people to litter and use plastics. Also, by having population data we are able to normalize our data and look at per capita measurements which will give us a better comparison between countries. We also believed that GDP would play an interesting role in our analysis. Typically, a higher GDP would mean that a country has more opportunities for consumption which could lead to more plastic waste. However, countries with higher GDP's are also more likely to have better waste managment systems that could reduce plastic waste.

```{r}
#renaming countries in my dataset to have ISO-2 codes that our API uses
plastics <- plastics %>% 
  filter(country != "EMPTY") %>% 
  mutate(iso2 = countrycode(country, 
                            origin = "country.name", 
                            destination = "iso2c"))

iso2_list <- unique(plastics$iso2)
iso2_list <- iso2_list[iso2_list != "TW"]
```

```{r}
#| eval: false
#testing API for one country gdp

us_url <- "https://api.worldbank.org/v2/country/US/indicator/NY.GDP.MKTP.CD?format=json"

us_data <- fromJSON(us_url)

us_gdp <- us_data[[2]]
```

```{r}
#| label: GDP Function
get_gdp <- function(iso2) {
  
  url <- glue("https://api.worldbank.org/v2/country/{iso2}/indicator/NY.GDP.MKTP.CD?format=json")
  
  my_data <- fromJSON(url)
  
  if(is.null(my_data[[2]])){
    return(NULL)
  } else{
    return(my_data[[2]] %>% 
             filter(date %in% c("2019","2020")) %>% 
             select(date, value) %>% 
             mutate(iso2c = iso2,
                    value = value/1e9) %>% 
             rename(gdp_billions = value))
  }
}

get_gdp("AU")

gdp_data <- map_dfr(iso2_list, get_gdp)
```

```{r}
#| eval: false
#Testing API for one country population

us_url <- "https://api.worldbank.org/v2/country/US/indicator/SP.POP.TOTL?format=json"

us_data <- fromJSON(us_url)

us_pop <- us_data[[2]]
```

```{r}
get_population <- function(iso2) {
  
  url <- glue("https://api.worldbank.org/v2/country/{iso2}/indicator/SP.POP.TOTL?format=json")
  
  my_data <- fromJSON(url)
  
  if(is.null(my_data[[2]])) {
    return(NULL)
  } else {
    return(my_data[[2]] %>% 
             filter(date %in% c("2019","2020")) %>%
             select(date, value) %>% 
             mutate(iso2c = iso2))
  }
}

get_population("AU")

population_data <- map_dfr(iso2_list, function(x) {
  Sys.sleep(1)
  get_population(x)
})
```

## External CO~2~ data:

We believed that another vital variable that may be related to the amount of plastic waste in countries is there CO~2~ emissions. We believe that CO~2~ emissions are likely correlated with higher industrial activity and energy use, which can also be linked to plastic waste. Because of this, the addition of CO~2~ emissions on our analysis can give us a better understanding of the underlying reason for larger plastic waste in certain countries.

```{r}
co2 <- readr::read_csv(
  "https://raw.githubusercontent.com/owid/co2-data/master/owid-co2-data.csv")
```

```{r}
#| label: Meta Data Creation

co2_sm <- co2 %>%
  filter(year %in% c(2019, 2020))

co2_sm <- co2_sm %>%
  mutate(iso2 = countrycode(iso_code, "iso3c", "iso2c"))

gdp_data <- gdp_data %>%
  mutate(date = as.numeric(date))

population_data <- population_data %>%
  mutate(date = as.numeric(date))

meta_set <- plastics %>%
  mutate(
    parent_company = ifelse(
      parent_company %in% c("NULL", "null"),
      "Grand Total",
      parent_company
    )
  ) %>%
  left_join(gdp_data, by = c("iso2" = "iso2c", "year" = "date")) %>%
  left_join(population_data, by = c("iso2" = "iso2c", "year" = "date")) %>%
  left_join(co2_sm, by = c("iso2", "year")) %>%
  mutate(
    gdp_per_capita = gdp_billions * 1e9 / population,
    plastic_waste_per_capita = grand_total / population
  )

meta_set_2019 <- meta_set %>% 
  filter(year == 2019)

meta_set_2020 <- meta_set %>% 
  filter(year == 2020) %>%
  group_by(country.x) %>%   
  summarise(
    grand_total = sum(grand_total, na.rm = TRUE),
    population = first(population),
    gdp_billions = first(gdp_billions),
    .groups = "drop"
  ) %>%
  mutate(parent_company = "Grand Total",
         year = 2020)
  
  
```

# Research Questions:

1.  What is the relationship between GDP per capita and plastic waste generation across countries in 2019-2020? Did the 2020 pandemic have the same effect on both?

2.  What is the relationship between GDP per capita and CO₂ emissions per capita across countries in 2019, and how does population size influence this relationship?

## Exploring the Link Between GDP per Capita and Plastic Waste:

GDP per capita provides a useful lens when examining the difference in plastic waste across countries. Initially it may seem that GDP is just used to measure the wealth of a country, but it also captures variations in income, consumption, and infrastructure, all factors that may influence both the amount of plastic waste generated and the steps that countries may take to reduce their waste. In this analysis, we investigate the relationship between GDP per capita and plastic waste generated across countries between 2019-2020.

To begin the analysis, we first looked into what plastic waste looked like per country. We started by finding the countries with the highest plastic waste in order to see if a pattern arrises. From the table we see that the countries with the highest plastic waste tend to belong to Africa or Southeast Asia. However, we do not believe that this is a regional pattern. We believe that these countries are likely going to be lower income countries that have less developed waste collection and management systems in place.

```{r}
#| label: Total Plastic Waste by Country

country_waste_summary <- function(plastics) {
  plastics |>
    mutate(country = str_to_lower(country)) |>
    filter(parent_company == "Grand Total", country != "empty") |>
    group_by(country) |>
    summarise(
      total_waste = sum(grand_total, na.rm = TRUE),
      .groups = "drop"
    ) |>
    arrange(desc(total_waste))
}

country_waste <- country_waste_summary(plastics)

country_waste |>
  slice_head(n=10) |>
  gt() |>
  cols_label(
    country = "Country",
    total_waste = "Total Plastic Waste"
  ) |>
  fmt_number(
    columns = total_waste,
    decimals = 0
  ) |>
  tab_caption(
    "Table 1: Waste by Country (2019-2020)"
    ) |>
  tab_header(
    title = md("**Total Plastic Waste Collected by Country**"),
    subtitle = "Total across 2019–2020 cleanup events"
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  )
```

To better understand how countries are addressing plastic waste, we also examine the distribution of volunteers involved in cleanup efforts. By normalizing total plastic waste by the number of volunteers, we create a measure of plastic waste per volunteer, which allows for more meaningful comparisons across countries of different sizes. This metric provides information on volunteer efforts and may highlight differences between the scale of the problem and the available human resources. The results suggest that some countries with lower total plastic waste still have high waste per volunteer. This gives us insight into how many of these countries may be struggling with reducing their waste due to their lower income or infrastructure.

```{r}
#| label: Plastic Waste per Volunteer by Country 

countries <- plastics |>
  filter(parent_company == "Grand Total",
         !is.na(country)) |>
  distinct(country) |>
  pull(country)

waste_per_volunteer_country <- map_dfr(countries, function(c) {
  
  plastics |>
    filter(parent_company == "Grand Total",
           country == c) |>
    mutate(c = str_to_lower(c)) |>
    summarise(
      country = c,
      total_waste = sum(grand_total, na.rm = TRUE),
      total_volunteers = sum(volunteers, na.rm = TRUE),
      waste_per_volunteer = total_waste / total_volunteers
    )
  
}) |>
  arrange(desc(waste_per_volunteer))

waste_per_volunteer_country |>
  slice_head(n=10) |>
  gt() |>
  cols_label(
    country = "Country",
    total_waste = "Total Waste",
    total_volunteers = "Total Volunteers",
    waste_per_volunteer = "Waste per Volunteer"
  ) |>
  fmt_number(
    columns = c(total_waste, total_volunteers, waste_per_volunteer),
    decimals = 0,
    use_seps = TRUE
  ) |>
  tab_header(
    title = md("**Plastic Waste per Volunteer Participation by Country**"),
    subtitle = "Computed using grand total across countries"
  ) |>
  tab_caption(
    "Table 2: Waste per volunteer participation across countries, constructed using iteration."
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  )
```

Now that we understand how plastic waste is distributed between countries and the impact of volunteers, we can look into the relationship between GDP per capita and plastic waste. Earlier, we discussed that GDP per capita may have a competing effect on plastic waste due to a larger GDP meaning more opportunity for consumption leading to large plastic waste, but also a larger GDP allows countries to allocate more funds towards reducing their waste. However, from the chart below we can see that there appears to actually be a negative linear relationship between GDP per capita and plastic waste. This means that as a countries GDP per capita increases, we would expect that country to produce less plastic waste.

```{r}
#| label: Comparison of GDP per capita and plastic waste per capita (2019)
#| warning: false
#| fig-cap: "Scatterplot of GDP per Capita vs. Plastic Waste Per Capita in the year 2019"
#| fig-alt: "Scatterplot analyzing the relationship between GDP per Capita and Plastic Waste per Capita in the year 2019. It shows a negative linear relationship between the variables."

<<<<<<< HEAD
graph1 <- meta_set %>%
  filter(year == 2019, parent_company == "Grand Total") %>%
=======

graph1 <- meta_set_2019 %>%
  filter(parent_company == "Grand Total")

graph1 <- graph1 %>%
>>>>>>> fca8987 (I finally fixed this stupid issue)
  mutate(
    gdp_per_capita = gdp_billions * 1e9 / population,
    plastic_waste_per_capita = grand_total / population
  )

ggplot(graph1, aes(x = gdp_per_capita, y = plastic_waste_per_capita)) +
  geom_point(alpha = 0.6, color = "#2A9D8F") +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "GDP per Capita vs Plastic Waste per Capita (2019)",
    x = "GDP per capita",
    y = "Plastic Waste per capita"
  ) +
  scale_x_log10() +
  scale_y_log10() +
  theme_ft_rc() +
  theme(
    plot.title = element_text(size = 16)
  )
```

While visualizations can give us an initial sense of the relationship between GDP per capita and plastic waste per capita, they are not sufficient enough to evaluate its significance. To better understand this relationship we apply a linear regression model, which allows us to estimate the direction and strength of our relationship. The resulting line of best fit, $\hat{plastic waste}=-0.6864(GDP per capita) - 1.9542$, provides us with an equation to estimate how plastic waste per capita changes with GDP per capita.

This equation suggest that there is a negative linear relationship between GDP per capita and plastic waste per capita. From this, we are able to deduce that as our GDP per capita increases, we expect to see less plastic waste per capita. Furthermore, our GDP per capita is associated with a p-value of 0.0271, providing us with evidence that GDP per capita is a significant predictor of plastic waste per capita in countries.

```{r}
#| label: Statistical Model

model_data <- meta_set %>%
  filter(year %in% c(2019, 2020), parent_company == "Grand Total") %>%
  mutate(
    gdp_per_capita = gdp_billions * 1e9 / population,
    plastic_waste_per_capita = grand_total / population,
    year = factor(year)
  )

model <- lm(
  log10(plastic_waste_per_capita) ~ log10(gdp_per_capita),
  data = model_data
)

model_table <- summary(model)$coefficients %>%
  as.data.frame() %>%
  tibble::rownames_to_column("term") %>%
  mutate(
    term = recode(
      term,
      `(Intercept)` = "Intercept",
      `log10(gdp_per_capita)` = "GDP Per Capita",
      `year2020` = "Year 2020"
    )
  )

model_table %>%
  gt() %>%
  cols_label(
    term = "Variable",
    Estimate = "Estimate",
    `Std. Error` = "Std. Error",
    `t value` = "t-value",
    `Pr(>|t|)` = "p-value"
  ) %>%
  fmt_number(
    columns = c(Estimate, `Std. Error`, `t value`, `Pr(>|t|)`),
    decimals = 4
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) %>%
  tab_header(
    title = md("**Regression Results: GDP, Year, and Plastic Waste**"),
    subtitle = md("Log-log linear model using 2019 and 2020 country-level data")
  ) %>%
  tab_caption(
    md("Table 3: Linear regression examining how GDP per capita and year relate to plastic waste per capita.")
  )
```

One of the key things to keep in mind with this analysis is the time that the data was collected. Our data is from 2019-2020 with 2020 being the start of the COVID-19 pandemic. Because of that, we thought we should look at the difference in the relationship between GDP per capita and plastic waste between those years. Using the bubble plot, we can see that in 2019 there was a steeper negative relationship than there was in 2020. We also generally see smaller values of plastic waste per capita in 2020, which is likely due to less consumption, less generation, and less people leaving their homes.

```{r}
#| label: Bubbleplot of GDP per Capita and Plastic Waste by Population Size Facted by Year
#| warning: false
#| fig-cap: "Relationship between GDP per capita and plastic waste per capita across countries in 2019 and 2020, with separate panels for each year."
#| fig-alt: "Faceted scatterplot showing GDP per capita versus plastic waste per capita for countries in 2019 and 2020. Each panel includes a fitted trend line. The relationship appears positive in both years, allowing comparison of how the pattern changes between 2019 and 2020."

graph2 <- meta_set_2019 %>%
  filter(parent_company == "Grand Total") %>%
  mutate(
    gdp_per_capita = gdp_billions * 1e9 / population,
    plastic_waste_per_capita = grand_total / population
  ) %>%
  filter(
    gdp_per_capita > 0,
    plastic_waste_per_capita > 0,
  ) %>%
  mutate(year = factor(year))     

graph3 <- meta_set_2020 %>%
  mutate(
    gdp_per_capita = gdp_billions * 1e9 / population,
    plastic_waste_per_capita = grand_total / population
  ) %>%
  filter(
    gdp_per_capita > 0,
    plastic_waste_per_capita > 0,
  ) %>%
  mutate(year = factor(year)) 

p1 <- ggplot(
  graph2,
  aes(
    x = gdp_per_capita,
    y = plastic_waste_per_capita,
    size = population
  )
) +
  geom_point(alpha = 0.6, color = "#2A9D8F") +
  geom_smooth(method = "lm", se = FALSE, show.legend = FALSE) +
  scale_size(range = c(1.5, 10), name = "Population") +
  scale_x_log10(labels = scales::label_dollar()) +
  scale_y_log10(labels = scales::label_scientific()) +
  theme_minimal() +
  guides(size = guide_legend(override.aes = list(shape = 16))) +
  theme_ft_rc() +
  labs(
    x = "GDP per Capita (USD)",
    y = "Plastic Waste per Capita",
    title = "GDP per Capita and Plastic Waste per Capita Across Countries",
    subtitle = "Comparison of 2019 and 2020"
  ) +
  theme(
    plot.title = element_text(size = 13),
    plot.subtitle = element_text(size = 11)
  )

p2 <- ggplot(
  graph3,
  aes(
    x = gdp_per_capita,
    y = plastic_waste_per_capita,
    size = population
  )
) +
  geom_point(alpha = 0.6, color = "#2A9D8F") +
  geom_smooth(method = "lm", se = FALSE, show.legend = FALSE) +
  scale_size(range = c(1.5, 10), name = "Population") +
  scale_x_log10(labels = scales::label_dollar()) +
  scale_y_log10(labels = scales::label_scientific()) +
  theme_minimal() +
  guides(size = guide_legend(override.aes = list(shape = 16))) +
  theme_ft_rc() +
  labs(
    x = "GDP per Capita (USD)",
    y = "Plastic Waste per Capita"
  ) +
  theme(
    plot.title = element_text(size = 13),
    plot.subtitle = element_text(size = 11)
  )



base_theme <- theme_ft_rc() +
  theme(
    plot.title = element_blank(),
    plot.subtitle = element_blank(),
    legend.position = "right",
    axis.text.x = element_text(angle = 30, hjust = 1),
    panel.grid.minor = element_blank()
  )

p1_clean <- p1 +
  labs(title = "2019", subtitle = NULL) +
  base_theme

p2_clean <- p2 +
  labs(title = "2020", subtitle = NULL) +
  base_theme

p1_clean + p2_clean +
  plot_layout(ncol = 2, guides = "collect") +
  plot_annotation(
    title = "GDP per Capita and Plastic Waste per Capita Across Countries",
    subtitle = "Comparison of 2019 and 2020"
  ) &
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 11),
    legend.position = "right"
  )
```

To truly understanding the impact of the pandemic, we look at the percent change in plastic waste from 2019 to 2020. If plastic waste per capita did decrease between the two years, we would expect to see a negative percent change. From the data, we find a percent change of -61.29%, indicating a significant decrease in plastic waste per capita over the two year span. This result supports our earlier observation that plastic waste per capita in 2020 was lower than in 2019. Therefore, our analysis suggests that during the pandemic there was a significant reduction in plastic waste per capita.

```{r}
#| label: Affect of COVID on Plastic Waste

plastic_data <- meta_set %>%
  filter(parent_company == "Grand Total") %>%
  group_by(year) %>%
  summarise(
    total_plastic = sum(grand_total),
    total_population = sum(population),
    per_capita = total_plastic / total_population
  )

plastic_change <- plastic_data %>%
  summarise(
    pct_change = (per_capita[year == 2020] - per_capita[year == 2019]) /
                 per_capita[year == 2019] * 100
  )

summary_table<- tibble(
  waste_metric = "Plastic per capita",
  percentage_change = plastic_change$pct_change
)

summary_table %>% 
  gt() %>% 
  tab_header(
    title = md("**Change in Plastic Waste (2019-2020)**"),
    subtitle = md("Percent change in per capita value")
  ) %>% 
  fmt_number(
    columns = percentage_change,
    decimals = 2
  ) %>% 
  cols_label(
    waste_metric = "Plastic Waste",
    percentage_change = "Percent Change (%)"
  ) %>% 
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) %>% 
  tab_caption(
    md("Table 4: Examining the percent change in plastic waste from 2019 to 2020")
  )
```

## Exploring the Link Between GDP per Capita and CO₂ Emissions per Capita:

With the introduction of GDP and CO~2~ emissions into our data set, we thought it would be interesting to explore the relationship between those variables. We assumed that for countries with higher GDP values, there are going to be more CO~2~ emissions due to the larger consumption and infrastructure. In order to see if this is impacted by population, we created a table comparing the per capita values with the countries population.

```{r}
#| label: GDP per Capita vs CO2 Emission per Capita in 2019

country_gdp_co2_summary <- function(data) {
  
  data |>
    filter(
      year == 2019,
      !is.na(country.x),
      country.x != "Taiwan_ Republic of China (ROC)"
    ) |>
    group_by(country.x) |>
    summarise(
      gdp_per_capita = mean(gdp_billions * 1e9 / population, na.rm = TRUE),
      co2_per_capita = mean(co2_per_capita, na.rm = TRUE),
      population = mean(population, na.rm = TRUE),
      .groups = "drop"
    ) |>
    arrange(desc(population))
}

gdp_co2_table <- country_gdp_co2_summary(meta_set)

gdp_co2_table |>
  slice_head(n = 10) |>
  gt() |>
  cols_label(
    country.x = "Country",
    gdp_per_capita = "GDP Per Capita",
    co2_per_capita = "CO2 per Capita",
    population = "Population") |>
  fmt_number(
    columns = c(gdp_per_capita, co2_per_capita, population),
    decimals = 0,
    use_seps = TRUE
  ) |>
  tab_header(
    title = md("**GDP vs CO2 Emissions per Capita (2019)**"),
    subtitle = "Top 10 countries sorted by population"
  ) |>
  tab_caption(
    "Table 5: Countries GDP per capita & CO2 per capita, sorted by population"
  ) |>
  fmt_currency(
  columns = contains("gdp"),
  currency = "USD",
  decimals = 0
)
```

To further analyze this relationship, we thought it would be a good idea to look at the per capita values compared to the countries global ranking in GDP and CO~2~ emissions. We arranged our table in descending order of GDP global ranking to see how their CO~2~ global ranking compares. It does appear that there is a relationship between the two variables as the GDP global ranking is usually relatively close to the CO~2~ global ranking. This further supports our claim from earlier predicting that countries with a higher GDP would have higher CO~2~ emissions.

```{r}
#| label: Comparing GDP per Capita and CO2 Emissions per Capita with Countries Global Ranking

countries <- meta_set |>
  filter(
    year == 2019,
    !is.na(country.x),
    country.x != "Taiwan"
  ) |>
  distinct(country.x) |>
  pull(country.x)

ranked_table <- map_dfr(countries, function(c) {
  meta_set |>
    filter(country.x == c, year == 2019) |>
    summarise(
      country = c,
      gdp_per_capita = mean(gdp_billions * 1e9 / population, na.rm = TRUE),
      co2_per_capita = mean(co2_per_capita, na.rm = TRUE)
    )
}) |>
  filter(!is.na(gdp_per_capita), !is.na(co2_per_capita)) |>
  mutate(
    gdp_rank = min_rank(desc(gdp_per_capita)),
    co2_rank = min_rank(desc(co2_per_capita))
  ) |>
  arrange(gdp_rank)

ranked_table |>
  slice_head(n = 10) |>
  gt() |>
  cols_label(
    country = "Country",
    gdp_per_capita = "GDP Per Capita",
    co2_per_capita = "CO2 Per Capita",
    gdp_rank = "GDP Rank",
    co2_rank = "CO2 Rank"
  ) |>
  fmt_currency(
    columns = contains("gdp"),
    currency = "USD",
    decimals = 0
  ) |>
  fmt_number(
    columns = c(co2_per_capita),
    decimals = 2,
    use_seps = TRUE
  ) |>
  fmt_number(
    columns = c(gdp_rank, co2_rank),
    decimals = 0
  ) |>
  tab_spanner(
    label = "Per-Capita Measures",
    columns = c(gdp_per_capita, co2_per_capita)
  ) |>
  tab_spanner(
    label = "Global Rankings",
    columns = c(gdp_rank, co2_rank)
  ) |>
  data_color(
    columns = c(gdp_rank, co2_rank),
    fn = scales::col_numeric(
      palette = c("#DCEAF7", "#2C7FB8"),
      domain = c(1, 30)
    )
  ) |>
  tab_header(
    title = md("**GDP and CO2 Per-Capita Ranking Comparison**"),
    subtitle = "Top 10 countries ranked by GDP per capita in 2019"
  ) |>
  tab_caption(
    "Table 6: Comparison of country rankings for GDP per capita and CO2 emissions per capita in 2019."
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  )
  
```

By just using tables, we are losing a lot of information that a visualization can provide. In order to compensate for that, we created a bubble plot of the relationship between GDP per capita and CO~2~ emissions per capita, with the size of the bubble correlating to the population size. From this graph there is a clear positive relationship between the variables. We see that as GDP per capita increases, so does CO~2~ emissions per capita. It also appears that countries with larger populations (signified by larger bubbles), also tend to have larger CO~2~ emissions than countries with smaller populations.

```{r}
#| warning: false
#| label: fig-rq2-bubble
#| fig-cap: "GDP per capita and CO2 emissions per capita across countries in 2019 and 2020, with bubble size representing population."
#| fig-alt: "Faceted bubble plot of GDP per capita versus CO2 emissions per capita for countries in 2019 and 2020. Bubble size represents population and color indicates GDP level. Countries with higher GDP per capita tend to also have higher CO2 emissions per capita."

model_data %>%
  arrange(desc(population)) %>% 
  mutate(
    country.x = as.factor(country.x),
    gdp_level = case_when(
      is.na(gdp_billions) ~ "Other",
      gdp_billions < 100 ~ "Low",
      gdp_billions < 500 ~ "Lower-Mid",
      gdp_billions < 1000 ~ "Upper-Mid",
      gdp_billions >= 1000 ~ "High"
    ),
    gdp_level = factor(
      gdp_level,
      levels = c("Low", "Lower-Mid", "Upper-Mid", "High", "Other")
    )
  ) %>% 
  ggplot(aes(
    x = gdp_per_capita,
    y = co2_per_capita,
    size = population,
    color = gdp_level
  )) +
  geom_point(alpha = 0.65) +
  facet_wrap(~year) +
  scale_size(range = c(2, 14), name = "Population") +
  scale_color_viridis_d(option = "A", end = 0.9, name = "GDP Level") +
  scale_x_log10(labels = scales::label_dollar()) +
  scale_y_log10(labels = scales::label_number()) +
  theme_ft_rc() +
  labs(
    x = "GDP Per Capita",
    y = "CO2 Emissions Per Capita",
    title = "Bubble Plot of GDP and CO2 Emissions Per Capita",
    subtitle = "Comparison of 2019 and 2020, with bubble size representing population"
  ) +
  theme(
    plot.title = element_text(size = 16),
    plot.subtitle = element_text(size = 12)
  )
```

Again, we were interested in whether the COVID-19 pandemic affected this relationship. In order to see the affect of the pandemic, we created two scatterplots analyzing the relationship between GDP per capita and CO~2~ emissions per capita by year. From the graph, we see that in 2019 and 2020, there is a strong positive relationship present. In 2020, this relationship does appear to be slightly less steep, likely due to the small decrease in CO~2~ emissions during the pandemic. However, there does not appear to be any indiciation that the pandemic had an affect on the relationship between GDP per capita and CO~2~ emissions per capita.

```{r}
#| label: Facted Scatterplot of GDP per Capita vs CO2 Emissions per Capita by Year
#| warning: false
#| fig-cap: "Scatterplot analyzing the relationship between GDP per capita vs CO2 emissions per capita across the years 2019-2020"
#| fig-alt: "Scatterplot that shows the relationship between GDP per capita vs CO2 emissions per capita, split by years 2019 and 2020. Both scatterplots show a positive linear relationships between the variables, with the relationship being almost identical between years."

graph3 <- meta_set %>%
  filter(parent_company == "Grand Total") %>%
  mutate(
    year = factor(year),
    gdp_per_capita = gdp_billions * 1e9 / population,
    plastic_waste_per_capita = grand_total / population
  ) %>%
  filter(
    gdp_per_capita > 0,
    co2_per_capita > 0
  )

ggplot(
<<<<<<< HEAD
  data = model_data,
=======
  data = graph3 %>%
    filter(year %in% c(2019, 2020), parent_company == "Grand Total"),
>>>>>>> fca8987 (I finally fixed this stupid issue)
  mapping = aes(
    x = gdp_per_capita,
    y = co2_per_capita
  )
) +
  geom_point(color = "#2C7FB8", alpha = 0.7, size = 2.5) +
  geom_smooth(method = "lm", se = FALSE, color = "#D95F0E", linewidth = 1) +
  scale_x_log10(labels = scales::label_dollar()) +
  scale_y_log10(labels = scales::label_number()) +
  facet_wrap(~year) +
  labs(
    title = "GDP per Capita vs CO2 Emissions per Capita",
    subtitle = "Comparison of country-level relationships in 2019 and 2020",
    x = "GDP per Capita",
    y = "CO2 Emissions per Capita"
  ) +
  theme_ft_rc()
```
